Nengo: A Python tool for building large-scale functional brain models

### 2013-10-04 nengo.tex
Edits from Chris's feedback, round 2

To describe an NEF model,
PyNengo defines six core objects.
\begin{enumerate}
  \item The \texttt{Model} encapsulates the entire model and
    exposes a simplified interface to other objects.
  \item The \texttt{Ensemble} contains a group of neurons
    that encodes a time-varying vector of real numbers.
  \item The \texttt{Node} represents non-neural information,
    such as sensory inputs.
  \item The \texttt{Connection} describes how
    nodes and ensembles are connected.
  \item The \texttt{Network} is a group of interconnected
    nodes and ensembles.
  \item The \texttt{Probe} gathers data during a simulation
    for later analysis.
\end{enumerate}

@@

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
When creating an ensemble,
the user must provide the dimensionality
(i.e., the length of the vector it represents),
and an object that describes
a population of neurons.
For example,
\begin{quote}
  \texttt{Ensemble('A', nengo.LIF(50, tau\_ref=0.002), 1)}
\end{quote}
describes an ensemble representing a one-dimensional vector
with 50 leaky integrate-and-fire neurons \citep{lapicque1907}
that have a refractory period of 2 milliseconds.
The \texttt{nengo.LIF} class defines
the parameters of the LIF neurons symbolically
so that each simulator can compute
the LIF nonlinearity efficiently.
The neuron model used by the ensemble
is changed by passing in a different symbolic neuron object;
however, the simulator used must be aware
of that type of neuron.

@@

A node represents information
that cannot be decoded from an ensemble.
As a simple example,
a node can be used to model sensory stimuli
that are predefined functions of time.
As a more sophisticated example,
a node can be used to model
a complex experimental environment that
both provides input to the neural model
and responds to the neural model's output.
Nodes allow many of the special-purpose
routines that are necessary in real-world
models to be integrated
as core components of a Nengo model.
This makes Nengo models more explicit,
and enables simulators
to control and optimize node execution.

@@

Networks can vary dramatically in size,
and can be composed of other networks.
The \texttt{Integrator} network, for example,
is composed of only one recurrently connected ensemble.
By encapsulating that logic in a network,
the purpose of that ensemble is made explicit.
On the other end of the spectrum,
the \texttt{BasalGanglia} network
is composed of five groups of ensembles
connected with several specific functions
that together implement a ``winner-take-all'' circuit.
Encapsulating the code to create those ensembles
and connections in a network subclass
makes a complicated section of code
easy to include in many different models.

@@

A \texttt{Probe} denotes that
a particular part of another object
will be recorded throughout a simulation.
Nengo models contain many variables
that change over time,
including membrane potentials,
spike events, and encoded vectors.
It is resource intensive to store the values of
large numbers of variables
at each timestep, and it is also not necessary,
as typically only a small fraction
of these variables are analyzed after a simulation.
The modeller chooses which variables to
record by connecting a probe to an object.

Like nodes, a probe could be implemented
outside of the model.
However, doing so requires detailed knowledge
of the simulator,
and can incur significant overhead
if not implemented carefully.
For these reasons, we have made probes
a core component of a Nengo model,
and are therefore explicit
and optimizable.
Further, since probes are described
at a symbolic level,
the underlying implementation
can output probed data in many different formats.
Currently, simulators store probed data
directly in memory,
but the ability to store data
in files or to stream data
directly to sockets will be added.

\begin{figure}
\begin{center}
  \includegraphics[width=\textwidth]{sim}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:sim} Figure \arabic{figure}.}{
   Detailed breakdown of the PyNengo reference simulator
   running a simple model for two timesteps.
   (A) Code describing the model being simulated.
   It consists of ensemble A projecting the sine of its
   encoded signal to ensemble B,
   which is recurrently connected.
   (B) Diagram depicting the model being simulated.
   (C) A detailed diagram of how the reference simulator
   organizes this model. Signals (blue) represent the values
   tracked in the simulation.
   Operators (red) represent the computations done on signals.
   Signals can be grouped as low-level neural signals
   that are used to compute the nonlinear functions
   underlying neuron models,
   and high-level NEF signals that are used to
   drive neurons and track the signals
   that the neurons are representing.
   The operators that implement the decoding
   and encoding steps map between
   the low-level neural signals
   and the high-level NEF signals.
   (D) The signals tracked at the low level
   can be interpreted as a model
   commonly seen in computational neuroscience literature;
   a population of leaky integrate-and-fire neurons
   is driven by some time-varying input current, $J(t)$.
   These neurons project to a population
   of recurrently connected neurons.
   The connection weights between the two populations,
   and from the second population to itself,
   can be computed by the NEF's transformation
   principle, bypassing the need for
   the high-level NEF signals
   used by the reference simulator
   for speed and data collection purposes.
   (E) The signals tracked at the high level
   can be interpreted as a dynamical system
   that can be analyzed with control theory.
   State variable A simply represents its input,
   and passes its state to a sine function
   which becomes the input to B.
   State variable B is a simple linear system
   that can be described with the typical
   $\dot{x}(t) = A x(t) + B u(t)$ equation.
   These dynamical systems can be simulated
   directly, without the use of spiking neurons,
   in order to quickly analyze system behavior.}
\end{figure}

@@

Many models in theoretical neuroscience
are based on attractor networks \citep{amit1992, deco2003}.
The NEF has been used in the past
to implement many different types of
attractor networks \citep{eliasmith2005}.
Figure~\ref{fig:lorenz} depicts
a Nengo implementation of the Lorenz chaotic attractor
with a single ensemble
composed of 2000 leaky integrate-and-fire neurons.
We have implemented the Lorenz attractor
in PyNN for benchmarking purposes
(code not shown).

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.8\textwidth]{cconv}
  \begin{minipage}{0.46\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
import nengo
from nengo.networks import CircularConvolution
model = nengo.Model("Circular convolution")
model.make_node("Input A", [-0.21, 0.5, 0.12, 0.06])
model.make_ensemble("A", nengo.LIF(512), 4)
model.connect("Input A", "A")
model.make_node("Input B", [-0.18, 0.28, 0.18, -0.52])
model.make_ensemble("B", nengo.LIF(512), 4)
model.connect("Input B", "B")
cconv = model.add(CircularConvolution("Circular Convolution",
    neurons=nengo.LIF(1032), dimensions=4))
model.connect("A", cconv.A)
model.connect("B", cconv.B)
model.make_ensemble("Result", nengo.LIF(512), 4)
model.connect(cconv, "Result")
model.probe("A", filter=0.02)
model.probe("B", filter=0.02)
model.probe("Result", filter=0.02)
sim = model.simulator()
sim.run(0.5)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.46\textwidth}
    \includegraphics[width=\textwidth]{cconv_res}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:cconv} Figure
   \arabic{figure}.}{
   Circular convolution implemented with Nengo.
   (A) Diagram depicting the model.
   The input vectors, A and B, represent four-dimensional vectors
   which are mapped onto six ensembles within the
   Circular Convolution network through
   complicated transformation matrices.
   Each ensemble within the network represents a
   two-dimensional vector. The product of the two dimensions
   is projected through another complicated transformation matrix
   to compute the final four-dimensional result.
   Note that the complicated parts of the model
   are contained within the network;
   the number of ensembles and the transform matrices shown
   are automatically generated by the network depending on
   the dimensionality of the input vectors.
   (B) Nengo code to build and simulate the model
   for 0.5 seconds.
   (C) The result of the simulation.
   Straight horizontal lines represent
   the target values that each ensemble
   should represent.
   Wavy lines represent the decoded values
   for each dimension represented by the
   A, B, and Result ensembles
   (top, middle and bottom panels, respectively).
   The ensembles represent the correct values,
   after a startup transient of less than 0.1 seconds.}
\end{figure}

A common function computed by Nengo models
is circular convolution.
Circular convolution is how
Spaun binds two concepts together.
Binding is central to solving many
of the high level cognitive tasks
that make Spaun unique.
Circular convolution
is best implemented in a two-layer network,
rather than in a single connection.
This two-layer organization is simplified
the \texttt{CircularConvolution} network subclass.
The complexity encapsulated in that network
can be seen in Figure~\ref{fig:cconv}.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.32\textwidth]{bench_cchannel}
  \includegraphics[width=0.32\textwidth]{bench_lorenz}
  \includegraphics[width=0.32\textwidth]{bench_cconv}
\end{center}
 \textbf{
     \refstepcounter{figure}\label{fig:benchmarks}
     Figure \arabic{figure}.}{
     Benchmark results for several simulators
     on the example models described
     in section~\ref{sec:examples}.
     In (A) and (B), all of the simulators
     except the PyNengo OpenCL simulator
     were run on an Intel Core i7-965.
     JavaNengo used all 4 cores of this processor;
     all other simulator used only 1 core.
     The PyNengo OpenCL simulator
     was run on an NVidia GTX280 GPU.
     (A) Benchmark results from simulating
     the communication channel for 10 simulated seconds
     at a 1 millisecond timestep.
     For all model sizes, PyNengo simulators
     are faster than JavaNengo,
     which is significantly faster than NEURON and NEST,
     which are significantly faster than Brian.
     The full Brian results are not shown;
     for the largest model, Brian takes $\sim$768 seconds.
     (B) Benchmark results from simulating
     the Lorenz attractor for 10 simulated seconds
     at a 1 millisecond timestep.
     For most model sizes,
     the results are the same as (A),
     except that NEURON is notably faster.
     However, when analyzed, the NEURON model
     emits almost no spikes after the first
     simulated second, which may be
     why the simulation is relatively fast.
     The full results for Brian and NEST are not shown;
     for the largest model, they take $\sim$1467
     and $\sim$601 seconds respectively.
     (C) Benchmark results from simulating
     circular convolution for 1 simulated second
     at a 1 millisecond timestep.
     For the blue lines, the simulator used
     was the PyNengo OpenCL simulator.
     The CPU used for JavaNengo and
     the PyNengo reference simulator
     was an Intel Core i7-3770;
     all 4 cores were used by JavaNengo,
     while PyNengo's reference simulator
     only used one core.
     For all model sizes,
     the OpenCL simulator is faster
     than the JavaNengo simulator,
     which is faster than the PyNengo
     reference simulator.
     The reference simulator was only run
     up to 50 dimensions.
     The full results for JavaNengo
     are not shown; for the largest model,
     it takes $\sim$45 seconds.
     }
\end{figure}

@@

The communication channel and Lorenz attractor
are small models that demonstrate
the principles of the NEF.
Their small size enables us to write
PyNN scripts that implement roughly
the same functionality
with Brian \citep{goodman2009}, NEURON \citep{hines2009},
NEST \citep{eppler2008}, and PCSim \citep{pecevski2009}.
We run each parameter set five times
on the same machine,
and plot the mean.
In most cases, the coefficient of variation
is well below 0.1, except for two
outliers with coefficients of 0.18 and 0.22,
indicating that these means are robust.
The results, shown in Figure~\ref{fig:benchmarks}A and B,
suggest that all versions of Nengo are significantly
faster than the simulators accessible
through PyNN, especially
as the size of models increase.
This is likely due to Nengo's
use of factorized weight matrices,
rather than storing and computing with
the entire weight matrix
on each timestep.
While NEST and NEURON were not
run on multiple cores using message passing,
the reference simulator of PyNengo
also only uses one CPU core.
The results also suggest that PyNengo's
simulators are faster than JavaNengo's simulator.

As a real-world example,
we have also benchmarked
the circular convolution model.
Circular convolution is an important test case,
as a significant portion of Spaun's
2.5 million neurons are used to
implement circular convolution.
In this case, only versions of Nengo
were tested.
Instead of running each simulation multiple times,
we instead run the simulator for 10 timesteps
in order to fill various levels of CPU or GPU cache,
and then run the simulator for 1000 more timesteps;
there is very little variance using this method.
As can be seen in Figure~\ref{fig:benchmarks}C,
for large models, the OpenCL simulator
performs much faster than JavaNengo;
in particular, a Radeon 7970 GPU performs 500-dimensional
circular convolution faster than real time,
and 200 times faster than JavaNengo.
Additionally, although both JavaNengo
and the OpenCL simulator on a CPU
use all available CPU cores,
PyNengo's OpenCL simulator is significantly faster.

@@

There are many other neural simulators
dedicated to building large-scale neural models;
however, Nengo is unique in being built
on a theoretical framework
that has been validated
through the Spaun model.

@@

As shown in Figure~\ref{fig:benchmarks},
Nengo is able to simulate many models
faster than each of PyNN's simulators.
This is, in part,
because Nengo stores the factors
of the connection weight matrix,
rather than storing the entire matrix.
However, PyNN is able to simulate
many detailed neuron models
and learning rules,
and has access to the wealth of models
already created for each of its
accessible simulators.
PyNN is therefore better suited to
simulate detailed neural models.

@@

By enabling outside contributions,
and providing a user-friendly API
and many example models,
we hope to foster a rich community of modellers
contributing networks, neuron types, learning rules,
and simulators to Nengo.
Internally, we will
implementing the JavaNengo use cases
not currently covered by PyNengo,
and create a graphical interface
to build models and
interactively inspect simulations.

@@

PyNengo is the next generation of Nengo.
Though the project is still young,
it can already simulate most models
that have been built using the NEF.
It does this with 11\% as many lines of code
as its predecessor,
and interacts seamlessly with
other scientific Python tools.
While the reference simulator
is simple and easy to understand,
the OpenCL simulator is extremely fast;
it can simulate a large circular convolution model
200 times faster than JavaNengo.
This makes the creation and simulation
of models that are many times larger than Spaun
tractable with current hardware.
These models will further test
the NEF as a theory of the brain;
PyNengo will make those models
accessible to anyone with a modern computer.

@@

Trevor Bekolay led development of the
PyNengo object model and scripting interface,
wrote the text of the paper,
and prepared all of the figures.
James Bergstra led development of
the PyNengo reference simulator and OpenCL simulator,
edited text, created an early version
of Figure~\ref{fig:sim},
and ran the benchmarks shown in
Figure~\ref{fig:benchmarks}C.
Eric Hunsberger contributed significantly
to PyNengo and both of its simulators, and edited text.
Travis DeWolf led development
of a Theano-backed
version of PyNengo that
identified issues with Theano,
and provided the base for
the version of PyNengo described in this paper.
Terrence C. Stewart contributed
to PyNengo and helped implement
the PyNN scripts used in
sections~\ref{sec:examples} and \ref{sec:benchmark}.
Daniel Rasmussen contributed
to PyNengo and the reference simulator,
and edited text.
Xuan Choo contributed to PyNengo
and the reference simulator.
Aaron Voelker ran the benchmarks
shown in Figure~\ref{fig:benchmarks}A and B.
Chris Eliasmith oversaw all development,
contributed to PyNengo,
and co-created the NEF with Charles Anderson.

@@

\paragraph{Funding\textcolon}
NSERC CRC, CFI, and OIT.

>>>>

To describe an NEF model,
PyNengo defines six core objects.
\begin{enumerate}
  \item The \texttt{Model} encapsulates the entire model and
    exposes a simplified interface to other objects.
  \item The \texttt{Ensemble} contains a group of neurons
    that encodes a time-varying vector of real numbers.
  \item The \texttt{Node} represents non-neural information,
    such as sensory inputs and motor outputs.
  \item The \texttt{Connection} describes how
    nodes and ensembles are connected.
  \item The \texttt{Network} encapsulates a functionally related
    group of interconnected nodes and ensembles.
  \item The \texttt{Probe} gathers data during a simulation
    for later analysis.
\end{enumerate}
These six objects contain symbolic information
about a Nengo model,
enabling a strict division between
model construction and simulation.
This allows a Nengo model
to be run on multiple simulators.

@@

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
When creating an ensemble,
the user must provide a name,
an object that describes
a population of neurons,
and the dimensionality
(i.e., the length of the vector it represents).
For example,
\begin{quote}
  \texttt{Ensemble('A', nengo.LIF(50, tau\_ref=0.002), 1)}
\end{quote}
describes an ensemble, `A',
that uses 50 leaky integrate-and-fire neurons \citep{lapicque1907}
with a 2 millisecond refractory period
to represent a one-dimensional vector.
The \texttt{nengo.LIF} class defines
the parameters of the LIF neurons symbolically
so that each simulator can compute
the LIF nonlinearity efficiently.
The neuron model used by the ensemble
is changed by passing in a different symbolic neuron object;
however, the simulator used must be aware
of that type of neuron.

@@

A node represents information
that cannot be decoded from an ensemble.
As a simple example,
a node can be used to model sensory stimuli
that are predefined functions of time.
As a more sophisticated example,
a node can be used to model
a complex experimental environment that
both provides input to the neural model
and responds to the neural model's output.
Nodes allow Nengo to represent
neural components,
the body that those components drive,
and the environment that body interacts with
in a single unified model.
This makes Nengo models more explicit,
and enables simulators
to control and optimize node execution.

@@

Networks can vary dramatically in size,
and can be composed of other networks.
As a simple example, the \texttt{Integrator} network
is composed of only one recurrently connected ensemble.
By encapsulating that logic in a network,
the purpose of that ensemble is made explicit.
As a complex example,
the \texttt{BasalGanglia} network
is composed of five groups of ensembles
connected with several specific functions
that together implement a ``winner-take-all'' circuit.
Encapsulating the code to create those ensembles
and connections in a network subclass
makes a complicated section of code
easy to include in many different models.

@@

A \texttt{Probe} monitors
a particular part of another object
that will be recorded throughout a simulation.
Nengo models contain many variables
that change over time,
including membrane potentials,
spike events, and encoded vectors.
It is resource intensive to store the values of
large numbers of variables
at each timestep, and it is also not necessary,
as typically only a small fraction
of these variables are analyzed after a simulation.
The modeller chooses which variables to
record by connecting a probe to an object.

Like nodes, a probe could be implemented
outside of the model.
However, doing so requires detailed knowledge
of the simulator,
and can incur significant overhead
if not implemented carefully.
For these reasons, we have made probes
a core component of a Nengo model,
and are therefore explicit
and optimizable.
Further, since probes are described
at a symbolic level,
the underlying implementation
can output probed data in many different formats.
Currently, simulators store probed data
directly in memory,
but the ability to store data
in files or to stream data
directly to sockets is forthcoming.

\begin{figure}
\begin{center}
  \includegraphics[width=\textwidth]{sim}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:sim} Figure \arabic{figure}.}{
   Detailed breakdown of the PyNengo reference simulator
   running a simple model for two timesteps.
   (A) Code describing the model being simulated.
   It consists of ensemble A projecting the sine of its
   encoded signal to ensemble B,
   which is recurrently connected.
   (B) Diagram depicting the model being simulated.
   (C) A detailed diagram of how the reference simulator
   organizes this model. Signals (blue) represent the values
   tracked in the simulation.
   Operators (red) represent the computations done on signals.
   Signals can be grouped as low-level neural signals
   that are used to compute the nonlinear functions
   underlying neuron models,
   and high-level NEF signals that are used to
   drive neurons and track the signals
   that the neurons are representing.
   The operators that implement the decoding
   and encoding steps map between
   the low-level neural signals
   and the high-level NEF signals.
   (D) The signals tracked at the low level
   can be interpreted as a model
   commonly seen in computational neuroscience literature;
   a population of leaky integrate-and-fire neurons
   is driven by some time-varying input current, $J(t)$.
   These neurons project to a population
   of recurrently connected neurons.
   The connection weights between the two populations,
   and from the second population to itself,
   can be computed by the NEF's transformation
   principle, bypassing the need for
   the high-level NEF signals
   used by the reference simulator
   for speed and data collection purposes.
   (E) The signals tracked at the high level
   can be interpreted as a dynamical system.
   State variable A simply represents its input,
   and passes its state to a sine function
   which becomes the input to B.
   State variable B is a simple linear system
   that can be described with the typical
   $\dot{x}(t) = A x(t) + B u(t)$ equation.
   These dynamical systems can be simulated
   directly, without the use of spiking neurons,
   in order to quickly analyze system behavior,
   if desired.}
\end{figure}

@@

Many models in theoretical neuroscience
are based on attractor networks \citep{amit1992, deco2003}.
The NEF has been used in the past
to implement many different types of
attractor networks \citep{eliasmith2005}.
Figure~\ref{fig:lorenz} depicts
a Nengo implementation of the Lorenz chaotic attractor
with a single ensemble
composed of 2000 leaky integrate-and-fire neurons.
We have implemented the Lorenz attractor
in PyNN for benchmarking purposes
(code not shown; the PyNN script is 100 lines long,
while the Nengo script [see Figure~\ref{fig:lorenz}
is 20 lines long).

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.8\textwidth]{cconv}
  \begin{minipage}{0.46\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
import nengo
from nengo.networks import CircularConvolution
model = nengo.Model("Circular convolution")
model.make_node("Input A", [-0.21, 0.5, 0.12, 0.06])
model.make_ensemble("A", nengo.LIF(512), 4)
model.connect("Input A", "A")
model.make_node("Input B", [-0.18, 0.28, 0.18, -0.52])
model.make_ensemble("B", nengo.LIF(512), 4)
model.connect("Input B", "B")
cconv = model.add(CircularConvolution("Circular Convolution",
    neurons=nengo.LIF(1032), dimensions=4))
model.connect("A", cconv.A)
model.connect("B", cconv.B)
model.make_ensemble("Result", nengo.LIF(512), 4)
model.connect(cconv, "Result")
model.probe("A", filter=0.02)
model.probe("B", filter=0.02)
model.probe("Result", filter=0.02)
sim = model.simulator()
sim.run(0.5)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.46\textwidth}
    \includegraphics[width=\textwidth]{cconv_res}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:cconv} Figure
   \arabic{figure}.}{
   Circular convolution implemented with Nengo.
   (A) Diagram depicting the model.
   The input vectors, A and B, represent four-dimensional vectors
   which are mapped onto six ensembles within the
   Circular Convolution network through
   complicated transformation matrices
   that implement a discrete Fourier transform.
   Each ensemble within the network represents a
   two-dimensional vector. The product of the two dimensions
   is projected through another complicated transformation matrix
   that implements the inverse discrete Fourier transform,
   computing the final four-dimensional result.
   Note that the complicated parts of the model
   are contained within the network;
   the number of ensembles and the transform matrices shown
   are automatically generated by the network depending on
   the dimensionality of the input vectors.
   (B) Nengo code to build and simulate the model
   for 0.5 seconds.
   (C) The result of the simulation.
   Straight horizontal lines represent
   the target values that each ensemble
   should represent.
   Wavy lines represent the decoded values
   for each dimension represented by the
   A, B, and Result ensembles
   (top, middle and bottom panels, respectively).
   The ensembles represent the correct values,
   after a startup transient of less than 0.1 seconds.}
\end{figure}

A common function computed by Nengo models
is circular convolution.
This function is a central part
of the Semantic Pointer Architecture \citep{eliasmith2013}
that uses the NEF to build
a variety of cognitive models including Spaun.
Circular convolution
is best implemented in a two-layer network,
rather than in a single connection.
This two-layer organization is simplified in
the \texttt{CircularConvolution} network subclass.
The complexity encapsulated in that network
can be seen in Figure~\ref{fig:cconv}.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.32\textwidth]{bench_cchannel}
  \includegraphics[width=0.32\textwidth]{bench_lorenz}
  \includegraphics[width=0.32\textwidth]{bench_cconv}
\end{center}
 \textbf{
     \refstepcounter{figure}\label{fig:benchmarks}
     Figure \arabic{figure}.}{
     Benchmark results for several simulators
     on the example models described
     in section~\ref{sec:examples}.
     In (A) and (B), all of the simulators
     except the PyNengo OpenCL simulator
     were run on an Intel Core i7-965.
     JavaNengo used all 4 cores of this processor;
     all other simulator used only 1 core.
     The PyNengo OpenCL simulator
     was run on an NVidia GTX280 GPU.
     (A) Benchmark results from simulating
     the communication channel for 10 simulated seconds
     at a 1 millisecond timestep.
     For all model sizes, PyNengo simulators
     are faster than JavaNengo,
     which is significantly faster than NEURON and NEST,
     which are significantly faster than Brian.
     The full Brian results are not shown;
     for the largest model, Brian takes $\sim$768 seconds.
     (B) Benchmark results from simulating
     the Lorenz attractor for 10 simulated seconds
     at a 1 millisecond timestep.
     For most model sizes,
     the results are the same as (A),
     except that NEURON is notably faster.
     The full results for Brian and NEST are not shown;
     for the largest model, they take $\sim$1467
     and $\sim$601 seconds respectively.
     (C) Benchmark results from simulating
     circular convolution for 1 simulated second
     at a 1 millisecond timestep.
     For the blue lines, the simulator used
     was the PyNengo OpenCL simulator.
     The CPU used for JavaNengo and
     the PyNengo reference simulator
     was an Intel Core i7-3770;
     all 4 cores were used by JavaNengo,
     while PyNengo's reference simulator
     only used one core.
     For all model sizes,
     the OpenCL simulator is faster
     than the JavaNengo simulator,
     which is faster than the PyNengo
     reference simulator.
     The reference simulator was only run
     up to 50 dimensions.
     The full results for JavaNengo
     are not shown; for the largest model,
     it takes $\sim$45 seconds.
     }
\end{figure}

@@

The communication channel and Lorenz attractor
are small models that demonstrate
the principles of the NEF.
Their small size enables us to write
PyNN scripts that implement roughly
the same functionality
with Brian \citep{goodman2009}, NEURON \citep{hines2009},
NEST \citep{eppler2008}, and PCSim \citep{pecevski2009}.
We ran each parameter set five times
on the same machine,
and plot the mean.
In most cases, the coefficient of variation
is well below 0.1, except for two
outliers with coefficients of 0.18 and 0.22,
overall indicating that the reported means are robust.
The results, shown in Figure~\ref{fig:benchmarks}A and B,
suggest that all versions of Nengo are significantly
faster than the simulators accessible
through PyNN, especially
as the size of models increases.
This is likely due to Nengo's
use of factorized weight matrices,
rather than storing and computing with
the entire weight matrix
on each timestep.
While NEST and NEURON were not
run on multiple cores using message passing,
the reference simulator of PyNengo
also only uses one CPU core.
The results also suggest that PyNengo's
simulators are faster than JavaNengo's simulator.

As a larger-scale example,
we have also benchmarked
the circular convolution model.
Circular convolution is an important test case,
as a significant portion of Spaun's
2.5 million neurons are used to
implement circular convolution.
In this case, only versions of Nengo
were tested.
Instead of running each simulation multiple times,
we instead run the simulator for 10 timesteps
in order to fill various levels of CPU or GPU cache,
and then run the simulator for 1000 more timesteps;
there is very little variance using this method.
As can be seen in Figure~\ref{fig:benchmarks}C,
for large models, the OpenCL simulator
performs much faster than JavaNengo;
in particular, a Radeon 7970 GPU performs
500-dimensional circular convolution
with about half a million neurons
faster than real time,
and 50 times faster than JavaNengo.
Additionally, although both JavaNengo
and the OpenCL simulator on a CPU
use all available CPU cores,
PyNengo's OpenCL simulator is significantly faster.

@@

There are many other neural simulators
dedicated to building large-scale neural models;
however, Nengo is unique in being built
on a theoretical framework
that has been validated
through the Spaun model and other past work.

@@

As shown in Figure~\ref{fig:benchmarks},
Nengo is able to simulate many models
faster than each of PyNN's simulators.
This is, in part,
because Nengo stores the factors
of the connection weight matrix,
rather than storing the entire matrix.
However, PyNN is able to simulate
many detailed neuron models
and learning rules,
and has access to the wealth of models
already created for each of its
accessible simulators.
Because Nengo is in an earlier development stage,
many of these detailed neuron models
and learning rules remain to be added.
PyNN is therefore currently better suited to
simulate detailed neural models.

@@

By enabling outside contributions,
and providing a user-friendly API
and many example models,
we hope to foster a rich community of modellers
contributing networks, neuron types, learning rules,
and simulators to Nengo.
Internally, we will be
implementing the JavaNengo use cases
not currently covered by PyNengo,
and create a graphical interface
to build models and
interactively inspect simulations.

@@

PyNengo is the next generation of Nengo.
Though the project is still young,
it can already simulate most models
that have been built using the NEF.
It does this with 11\% as many lines of code
as its predecessor,
and interacts seamlessly with
other scientific Python tools.
While the reference simulator
is simple and easy to understand,
the OpenCL simulator is extremely fast;
it can simulate circular convolution models
50-200 times faster than JavaNengo,
which is faster than alternative simulators
on simpler models.
This makes the creation and simulation
of models that are many times larger than Spaun
tractable with current hardware.
These models will further test
the NEF as a theory of the brain;
PyNengo will make those models
accessible to anyone with a modern computer.

@@

Trevor Bekolay led development of the
PyNengo object model and scripting interface,
wrote the text of the paper,
and prepared all of the figures.
James Bergstra led development of
the PyNengo reference simulator and OpenCL simulator,
edited text, created an early version
of Figure~\ref{fig:sim},
and ran the benchmarks shown in
Figure~\ref{fig:benchmarks}C.
Eric Hunsberger contributed significantly
to PyNengo and both of its simulators, and edited text.
Travis DeWolf led development
of a Theano-backed
version of PyNengo that
identified issues with Theano,
and provided the base for
the version of PyNengo described in this paper.
Terrence C. Stewart contributed
to PyNengo, helped implement
the PyNN scripts used in
sections~\ref{sec:examples} and \ref{sec:benchmark},
and implemented the JavaNengo scripting interface
on which PyNengo is based.
Daniel Rasmussen contributed
to PyNengo and the reference simulator,
and edited text.
Xuan Choo contributed to PyNengo
and the reference simulator.
Aaron Voelker ran the benchmarks
shown in Figure~\ref{fig:benchmarks}A and B.
Chris Eliasmith oversaw all development,
contributed to PyNengo,
and co-created the NEF with Charles Anderson.

@@

\paragraph{Funding\textcolon}
NSERC Discovery, NSERC Graduate Fellowships,
NSERC Banting Fellowship, ONR (N000141310419)
and AFOSR (FA8655-13-1-3084).
